---
title: "A04_Documentation"
author: "Thompson Santiago"
date: "2026-06-01"
output: html_document
---

**__Understanding personal formatting quirks__**

*Single hashtag comments (#) refer to instructions from Sir Christian and are copy-pasted for the most part.*

*Double hashtag comments (##) are personal comments and usually attempt to explain the code or provide interpretations gleaned from its output.*

# **__Initialization__**

## Reading the file and installing the libraries needed
```{r init1, echo = TRUE, warning = TRUE}
# Read in the CSV file as a dataframe.
htypes <- data.frame(read.csv(file = "C:/DLSU/25-26/[4] LBBBI27/activities/a04/habitat_types.csv"))

# Install tidveryse and load.
library(tidyverse)

# Show the data structure using the tidyverse equivalent of the “str” base R.
str(htypes)
```

# Data shaping

## Subsetting for our purposes
```{r ds1, echo = TRUE, warning = TRUE}
# Subset the data by taking only the following columns: Transect, Hatypes, SmallTrees, and FloweringTrees.
subset_data <- select(htypes, "Transect", "Htypes", "SmallTrees", "FloweringTrees")
```

## Sort for small tree count
```{r ds2, echo = TRUE, warning = TRUE}
# From the subset of data, filter the habitat type “CVT” and sort the “SmallTrees” variable in descending order.
# Which transect has the most count of small trees, and how many?
cvt_desc_data <- filter(subset_data, Htypes=="CVT") |> arrange(desc(SmallTrees))
```

## Get mean of flowering tree number per habitat
```{r ds3, echo = TRUE, warning = TRUE}
# Also, calculate the mean number of flowering trees per habitat type.
# Which habitat types do not differ significantly in the mean number of flowering trees?
# fm = flower mean
cvt_fm <- filter(subset_data, Htypes=="ESG") |> summarize(mean(FloweringTrees))
esg_fm <- filter(subset_data, Htypes=="ESG") |> summarize(mean(FloweringTrees))
asg_fm <- filter(subset_data, Htypes=="ASG") |> summarize(mean(FloweringTrees))
## CVT and ESG do not significantly differ in flowering tree mean.
```

## Calculate difference between herb and ground cover.
```{r ds4, echo = TRUE, warning = TRUE}
# In your subset of data (e.g., from step #4), add a new column showing the difference between HerbsCover and GroundCover.
# Name the new column as “HerbGroundDiff.” Make sure to assign the output of this step to a new variable called “sel_df_mod.”
self_df_mod <- subset_data |> mutate(HerbGroundDiff=htypes$HerbsCover-htypes$GroundCover)

# Are there any transects with no difference? Can you identify at least three transects?
## IT1P5, IT3P1, ST1P3
```

## "lattice-ify" the table.
```{r ds5, echo = TRUE, warning = TRUE}
# Convert your self_df_mod output to a wider version (i.e., using pivot_wider).
# The output should have columns like:
# SmallTrees_ESG, FloweringTrees_ESG, HerbGroundDiff_ESG, SmallTrees_CVT, FloweringTrees_CVT, etc.
self_df_wide <- self_df_mod |> pivot_wider(names_from = Htypes, values_from = c(SmallTrees, FloweringTrees, HerbGroundDiff))
```

# Reflection
* This part is not code.*
```{r ref1, echo = TRUE, warning = TRUE}
# Tidyverse is nice in that the syntax is more streamlined and uses more words which I'm more used to, I think.
# The pipe also helps in not making this(kind(of(thing(happen)))).
```