---
title: "Santiago Linear Regression"
author: "Thompson Santiago"
date: "2026-07-06"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set-up Phase
*Libraries to be used, I don't think anything special for this one, also trying something new and that is to code directly in the .rmd file (used to copy-paste everything over but found that .rmd has some nice UI inclusions)*
```{r libraries}
library(tidyverse)
library(terra)
```

# A. Simple Linear Regression
*For this portion, I decided on acting out of question order after giving it a read through as it was essentially just one straight motion.*
```{r maindata}
# 2. Read in the species data and remove unnecessary rows with NAs.
herp_csv <- read.csv("C:/DLSU/25-26/[4] LBBBI27/activities/a03/herpetozoa-033-095-s001.csv")
herp_data <- herp_csv[complete.cases(herp_csv),]

# 4.1 From the species data, subset the longitude and latitude coordinates.
herp_spat <- herp_data[,c(2,3)]
class(herp_spat)

# 4.2 Then convert the coordinates to a spatial vector using the “vect” function from the “terra” package.
herp_vect <- vect(herp_spat, geom=c("Long","Lat"))
```
*This code block is dedicated to read-ins and establishment of supplementary data and data variables.*
```{r suppdata}
# 3. Read in the elevation data (ph_strm_geo.tif).
elev_data <- rast("C:/DLSU/25-26/[4] LBBBI27/activities/a10/ph_ strm_geo.tif")

# 5. Using the converted coordinates, extract elevation data for each point with the “extract” function in the “terra” package.
# (The extracted data will have two columns: “ID” and “Band_1.” The “Band_1” column contains the elevation.)
elev_extr <- extract(elev_data, herp_vect)

# 6.1 Subset the species data to retain only the species richness column (SPPTotal).
herp_rich <- matrix(herp_data[,6])

# 6.2 Then combine this column with the extracted elevation. Rename “Band_1” to “elevation.
herp_elev <- cbind(elev_extr, herp_rich)
herp_elev <- rename(herp_elev, "Elevation" = "Band_1")
```

*Using the fruits of the previous code locks to do number stuff! Single linear regression for now for part A.*
*In single quotes, my thoughts. In comments, the assignment instructions.*
```{r slr_nums}
# 7.1 Now, perform simple linear regression using the code from the pre-lab lecture,
'Raw values assigned to variable names for convenience'
slr_x <- herp_elev[,2]
slr_y <- herp_elev[,3]

xb <- mean(slr_x)
yb <- mean(slr_y)

'Secondary-level variables'
b <- sum((slr_x - xb) * (slr_y - yb)) / sum((slr_x - xb)^2)
a <- yb - b*xb

'Write the equation (I guess this will help with understanding what\'s going on'
cat("Regression Equation:\n")
cat("y =", round(a,3), "+", round(b,3), "x\n")

'Looks about right, setting it'
yhat <- a + b * slr_x

slr_data <- data.frame(
  "Elevation" = slr_x,
  "Species Richness" = slr_y,
  "Predicted" = round(yhat, 2)
)
```
*Who up plotting?*
```{r slr_plot}
# 7.2 including a plot. 
plot(herp_elev$"Elevation", herp_elev$"herp_rich",
     pch = 20,
     col = "darkcyan",
     xlab = "Elevation (m)",
     ylab = "Species Richness (#)",
     main = "Simple Linear Regression") +
abline(a, b,
       col = "darkred",
       lwd = 2)

# 7.3 What can you infer about the relationship between species and elevation?
'The single linear regression model observes and predicts decreases in species richness as elevation increases.'
```

# B. Multiple Linear Regression
*Read-ins of x-vars beyond the first.*
```{r precitemp}
# 1. Read in both annual precipitation and temperature data and extract the data following the instructions from A4–5.
ph_temps <- rast("C:/DLSU/25-26/[4] LBBBI27/activities/a03/annual_temperature.tif")
temp_extr <- extract(ph_temps, herp_vect)

ph_preci <- rast("C:/DLSU/25-26/[4] LBBBI27/activities/a03/annual_precipitation.tif")
preci_extr <- extract(ph_preci, herp_vect)

```

*C-Bind into one dataframe.*
```{r datacombi}
# 2. Combine the extracted data with the species and elevation data into a single dataframe.
herp_multi <- cbind(herp_elev$Elevation, temp_extr$annual_temperature, preci_extr$annual_precipitation)
herp_multi <- data.frame(herp_multi)
herp_multi <- rename(herp_multi, "Elevation" = "X1", "Annual Temperature" = "X2", "Annual Precipitation" = "X3")
class(herp_multi)
```
*Multiple linear regression*
```{r }
# 3. Perform multiple linear regression using the “lm” function.
m_model <- lm(herp_elev$"herp_rich" ~ herp_multi$"Elevation" + herp_multi$"Annual Temperature" + herp_multi$"Annual Precipitation")

coef(m_model)
ma <- coef(m_model)[1]
m1 <- coef(m_model)[2]
m2 <- coef(m_model)[3]
m3 <- coef(m_model)[4]

cat("y =", round(ma,3), "+", round(m1,3), "x1", "+", round(m2,3), "x2", "+", round(m3,3), "x3")

m_yhat <- predict(m_model) |> data.frame()
print(m_yhat)

summary(m_model)

# 4. Which relationship is statistically significant?
'Based on the summary of m_model (var name for multiple linear regression model), the annual temperature and annual precipitation of a location is significantly related to its species richness. This is not true of elevation of the place.'
```

*Okay that's all I have. I still don't know how to generate maps so I will have to submit a new version after all my technical issues are resolved and I learn how to use what I've not been able to learn so far.*